---
title: "EDA - BarrettE"
execute: 
  echo: true      
code-fold: false  
---

For my exploratory data analysis,  I've included some data visualizations that we decided not to include in the final project because they were not as relevant to our research questions/data storytelling. 

This includes a visualization meant to compare 2019 to 2024 directly. We decided not to include this because it didn't contain any interesting information for our story. 

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

```{r}
# Comparing 2019 vs. 2024
BookIns_2019 <- read_csv("../../data/ICE Initial Book-Ins by Arresting Agency and Month_ FY2019.csv")
BookIns_2024 <- read_csv("../../data/ICE Initial Book-Ins by Arresting Agency and Month_ EOFY2024.csv")
```


```{r}
BookIns_2019 <- BookIns_2019 |>
  mutate(Year = "2019")

BookIns_2024 <- BookIns_2024 |>
  mutate(Year = "2024")
```

```{r}
BookIns_2019_long <- BookIns_2019 |>
  pivot_longer(
    cols = -c(Agency, Year),
    names_to = "Month",
    values_to = "Count"
  )

BookIns_2024_long <- BookIns_2024 |>
  pivot_longer(
    cols = -c(Agency, Year),
    names_to = "Month",
    values_to = "Count"
  )

names(BookIns_2019_long)
names(BookIns_2024_long)
```



```{r}
BookIns_Comp <- bind_rows(BookIns_2019_long, BookIns_2024_long)

BookIns_Comp |> 
  ggplot(aes(x = Month, y = Count, fill = Agency)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Year) +
  labs(title = "Agency Book-Ins by Month",
       x = "Month",
       y = "Count") +
  theme_minimal()
```

```{r}
BookIns_Comp |> 
  filter(Month == "Total") |>
  ggplot(aes(x = Month, y = Count, fill = Agency)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Year) +                # Create facets for each Year
  labs(title = "Agency Book-Ins by Month",
       x = "Month",
       y = "Count") +
  theme_minimal()
```


```{r}
BookIns_Comp |> 
  filter(Month != "Total") |>
  ggplot(aes(x = Month, y = Count, fill = Agency)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Year) +                # Create facets for each Year
  labs(title = "Agency Book-Ins by Month",
       x = "Month",
       y = "Count") +
  theme_minimal()
```

```{r}
# Ensuring all columns except 'Agency' are numeric or character before pivoting
BookIns_2019 <- BookIns_2019 |>
  mutate(across(-Agency, as.character)) |>  # Convert all columns except 'Agency' to character
  pivot_longer(
    cols = -Agency,           # Pivot all columns except Agency
    names_to = "Month",       # Convert column names (months) into a new "Month" column
    values_to = "Count"       # Values will go into the "Count" column
  ) |>
  mutate(Year = "2019")        # Add the Year column to differentiate between datasets

# Same operation for BookIns_2024
BookIns_2024 <- BookIns_2024 |>
  mutate(across(-Agency, as.character)) |>  # Convert all columns except 'Agency' to character
  pivot_longer(
    cols = -Agency,           # Pivot all columns except Agency
    names_to = "Month",       # Convert column names (months) into a new "Month" column
    values_to = "Count"       # Values will go into the "Count" column
  ) |>
  mutate(Year = "2024")        # Add the Year column to differentiate between datasets

# Merging the datasets using a full join
BookIns_Comp <- BookIns_2019 |>
  full_join(BookIns_2024, by = c("Agency", "Month", "Count", "Year"))

# Reorder the months factor level to ensure they are in chronological order
month_order <- c("January", "February", "March", "April", "May", "June", 
                 "July", "August", "September", "October", "November", "December")

# Plotting BookIns_Comp with the months in the correct order and faceting by Year
BookIns_Comp |> 
  mutate(Month = factor(Month, levels = month_order)) |>
  ggplot(aes(x = Month, y = Count, fill = Agency)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Year) +                # Create facets for each Year
  labs(title = "Agency Book-Ins by Month",
       x = "Month",
       y = "Count") +
  theme_minimal()

# Filtering out "Total" values from the dataset and plotting again
BookIns_Comp_clean <- BookIns_Comp |> 
  filter(Month != "Total") |>  # Remove "Total" if it's present
  mutate(Month = factor(Month, levels = month_order))  # Reorder the months again

BookIns_Comp_clean |> 
  ggplot(aes(x = Month, y = Count, fill = Agency)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Year) +                # Create facets for each Year
  labs(title = "Agency Book-Ins by Month",
       x = "Month",
       y = "Count") +
  theme_minimal()

```

